A full approximation scheme multilevel method \\ for nonlinear variational inequalities

### 2023-07-23 paper/fascd.tex
More edits

Here we extend Tai's multilevel CD method by replacing the coarser-level corrections with a full approximation scheme (FAS; \cite{Brandt1977,Bruneetal2015}) approach.  Specifically, our \emph{full approximation scheme constraint decomposition} (FASCD) method extends the \cite{GraeserKornhuber2009,Tai2003} algorithms in the following directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems. The needed computations require neither an objective function nor linearity of the residual operator.
\item Box constraints (i.e.~both upper and lower obstacles) are allowed.
\item As with the original CD method \cite{Tai2003}, but unlike some implementations \cite[for example]{GraeserKornhuber2009}, and semismooth methods generally \cite{BensonMunson2006,Ulbrich2011}, the method can be applied even if the residual operator is defined only on those functions which satisfy the constraints.
\pef{I confess I find this explicit calling out of other algorithms a bit uncomfortable/too much. I wonder if the following is better?}
\item The method is strictly admissible, and so can be applied even if the residual operator is defined only on those functions which satisfy the constraints.
\item V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle}) are implemented.
\item We identify a ``telescoping'' subset property which can be exploited on the up-smoothing side of a V-cycle for greater performance.
\item When used in conjunction with a reduced-space active set method as a smoother, we demonstrate that F-cycle FASCD is an optimal solver for the different VI examples in Section \ref{sec:results}.  The work of the smoother is proportional to the number of unknowns on a given level because the number of Newton and Krylov iterations is fixed.
\end{enumerate}

Prior multilevel algorithms for VIs have various substantial limitations compared to the FASCD framework.  The projected full approximation scheme (PFAS) algorithm of Brandt \& Cryer \cite{BrandtCryer1983} is apparently limited to linear complementarity problems and projected (truncated) Gauss--Seidel smoothers.  While FASCD reduces to general, nonlinear FAS for PDEs when the inequality constraints are removed (Appendix \ref{app:reductions}), PFAS reduces to linear multigrid.  The multilevel CD iteration of Tai \cite{Tai2003} is formulated in terms of an objective function, so it only applies to optimization-type problems, and furthermore each of its coarser-level problems refers to the finest-level discretization.  The latter property is improved by reformulation using defect constraints and monotone injection operators (see \cite[Algorithm 4.7]{GraeserKornhuber2009}, or Section \ref{sec:cdmultilevel} herein).  The monotone multigrid methods of Kornhuber and co-authors \cite{GraeserKornhuber2009,Kornhuber1994} appear to be formulated only with Gauss--Seidel type smoothers, and applied only to optimization-type VI problems \cite[for example]{JouvetGraeser2013}.  Furthermore, monotone multigrid methods are more difficult to implement in FE libraries because modifications of the FE basis functions are needed to approximate the coarser-level coincidence sets.  \pef{My current understanding is that this refers to \emph{truncated} monotone multigrid; monotone multigrid just means using monotone injection operators appropriately. Let me double-check. And my understanding is that the `modification' of the basis functions is just zeroing the basis functions for vertices in the active set, so you don't have to modify the basis functions, just zero some rows/columns of the matrix.} No such invasive basis modifications are used either in Tai's CD, nor in our Firedrake library \cite{Rathgeberetal2016} implementation of FASCD.  However, our implementation exceeds monotone multigrid performance, including on a problem used to demonstrate the benefits of that method (Example \ref{ex:results:classical}; compare \cite[problem 7.1.1]{GraeserKornhuber2009}).

Regarding its limitations, FASCD solves only bound-constrained VI problems.  The method differs from the original FAS method only if the constraint set is proper (Appendix \ref{app:reductions}).

@@

To construct box constraints on all meshes in the hierarchy, let $\tilde{\mathcal{V}}^j$ denote the set of functions on the vertices of $\mathcal{T}^j$ with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  The finest-level \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ are constructed by suitable interpolation of $\underline{\gamma}$, $\overline{\gamma}$ (e.g.~by applying the Cl\'ement interpolation operator \elb{cite?}).  They must satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J. \label{eq:fe:boxconstraintrequirements}
\end{equation}
Analogously, we construct the discrete boundary data $g_D^J$ so that $\underline{\gamma}^J \le g_D^J \le \overline{\gamma}^J$ on those vertices of $\mathcal{T}^J$ which lie in $\partial_D \Omega$.  Then we define
\begin{equation}
\mathcal{K}^J = \left\{w\,:\,\underline{\gamma}^J \le w \le \overline{\gamma}^J \text{ on vertices of } \mathcal{T}^J, \, \text{ and } \, w|_{\partial_D\Omega} = g_D^J|_{\partial_D\Omega}\right\} \subset \mathcal{V}^J, \label{eq:fe:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of admissible FE functions. This is a nonempty, closed, and convex subset of $\mathcal{V}^J$.

@@

The monotone injections $\maxR,\minR$ \cite{GraeserKornhuber2009}, which will only be applied to constraints in $\tilde{\mathcal{V}}^j$, are nonlinear maps which maximize and minimize, respectively, nodal values over the open supports of coarser-mesh basis functions (i.e.~over the stars of the coarse vertices), Figure \ref{fig:Rplusminus}. These operators are constructed to enforce the property
\begin{equation}
\minR z \le z \le \maxR z,  \label{eq:fe:monotoneproperty}
\end{equation}
which does not hold for the standard injection operator $\iR$.
To make the construction precise, let $x_q^j$ denote the vertices of mesh $\mathcal{T}^j$, with corresponding basis functions $\psi_q^j(x)$.  For a $\tilde\RR$-valued nodal function $z\in\tilde{\mathcal{V}}^j$ we define functions in $\tilde{\mathcal{V}}^{j-1}$:
\begin{align}
(\maxR z)(x_p^{j-1}) &= \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}, \label{eq:fe:monotoneinjections} \\
(\minR z)(x_p^{j-1}) &= \min \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}. \notag
\end{align}
For standard refinement a coarse-mesh node $x_p^{j-1}$ is in the set of fine vertices $x_q^j$ such that $\psi_p^{j-1}(x_q^j)>0$, so \eqref{eq:fe:monotoneproperty} follows.  In the next Section we use $\maxR,\minR$ to construct ``level defect constraints'' such that the zero function is always admissible.  This construction depends on an additional ordering property of the monotone injections:
\begin{equation}
z\ge 0 \implies \minR z \ge 0, \qquad z \le 0 \implies \maxR z \le 0. \label{eq:fe:monotoneadditional}
\end{equation}

@@

\begin{example} \label{ex:results:classical}
Consider the unilateral 2D Laplacian obstacle problem, Example \ref{ex:plaplacian} with $p=2$ and admissible set $\mathcal{K} = \{v \ge \psi\} \subset W^{1,2}(\Omega)$.  We consider two problems, both over square domains $\Omega$, with the very different coincidence (active) sets $\{x\in\Omega \,:\, u(x)=\psi(x)\}$ shown in Figure \ref{fig:results:classical}.  The standard ``ball'' problem has an upper hemisphere obstacle $\psi$ \cite[Chapter 12]{Bueler2021}.  For the ``spiral'' problem \cite[problem 7.1.1]{GraeserKornhuber2009}, $\psi$ is constructed to produce the coincidence set shown.  For this experiment we used a coarsest regular-triangulation mesh with $m_0=5^2$ vertices (32 triangles).  The finest mesh, built by standard refinement, had $m_9=2049^2 \approx 4.2 \times 10^{6}$ vertices.  As our smoother we applied a single RS Newton iteration (Section \ref{sec:implementation}), with three (fixed) IC-preconditioned CG iterations to approximately solve the arising linear systems.  The coarse solver was RS$+$LU, iterated to convergence.  Tolerances \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$ were used in stopping criterion \eqref{eq:stoppingcriterion}.  The initial iterate was set to $w=\max\{0,\psi\}$.

@@

The 2D solver applied $P_1$ elements and a single processor, but in 3D the elements were chosen to be $Q_1$ (hexahedra) and the run parallel ($P=16$ processes).  A coarse mesh with $m_0=16^d$ vertices minimally-resolved the coincidence set.  The smoother was two (fixed) RS Newton iterations, with three preconditioned GMRES iterations to approximately solve the arising linear systems; the preconditioning was additive Schwarz method (ASM) with incomplete factorization (ILU) on each process.  The coarse solver was RS Newton and LU, iterated to convergence.  Tolerances \id{rtol} $=10^{-5}$ and \id{atol} $=10^{-9}$ were used in \eqref{eq:stoppingcriterion}.

@@

In contrast with monotone multigrid \cite{Kornhuber1994}, FASCD does not modify the FE discretization, but the possibility of acceleration via basis-level manipulations is a topic for future research.  On the other hand,  the convergence proofs in \cite{GraeserKornhuber2009,Kornhuber1994}, which appear to depend on particular smoother and basis-level manipulation choices, have no analog here.

@@

With these modifications, Algorithm \ref{alg:fascd} becomes the FAS V-cycle below, as a solver for \eqref{eq:app:fas:pde}.  It appears in various literature, for example \cite[Algorithm 14]{Bruneetal2015} and \cite[section 5.3.4]{Trottenbergetal2001}.

>>>>

Here we extend Tai's multilevel CD method by replacing the coarser-level corrections with a full approximation scheme (FAS; \cite{Brandt1977,Bruneetal2015}) approach.  Specifically, our \emph{full approximation scheme constraint decomposition} (FASCD) method extends the \cite{GraeserKornhuber2009,Tai2003} algorithms in the following directions:
\pef{These aren't all really extensions. Maybe we could say `enjoys the following advantages'? We distinguish it from prior work below, so this section does not need to do that.}
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems. The needed computations require neither an objective function nor linearity of the residual operator.
\item Box constraints (i.e.~both upper and lower obstacles) are allowed.
\item As with the original CD method \cite{Tai2003}, but unlike some implementations \cite[for example]{GraeserKornhuber2009}, and semismooth methods generally \cite{BensonMunson2006,Ulbrich2011}, the method can be applied even if the residual operator is defined only on those functions which satisfy the constraints.
\pef{I confess I find this explicit calling out of other algorithms a bit uncomfortable/too much. I wonder if the following is better?}
\item The method is strictly admissible, and so can be applied even if the residual operator is defined only on those functions which satisfy the constraints.
\item V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle}) are implemented.
\item We identify a ``telescoping'' subset property which can be exploited on the up-smoothing side of a V-cycle for greater performance.
\item When used in conjunction with a reduced-space active set method as a smoother, we demonstrate that F-cycle FASCD is an optimal solver for the different VI examples in Section \ref{sec:results}.  The work of the smoother is proportional to the number of unknowns on a given level because the number of Newton and Krylov iterations is fixed.
\end{enumerate}

Prior multilevel algorithms for VIs have various substantial limitations compared to the FASCD framework.  The projected full approximation scheme (PFAS) algorithm of Brandt \& Cryer \cite{BrandtCryer1983} is apparently limited to linear complementarity problems and projected (truncated) Gauss--Seidel smoothers.  While FASCD reduces to general, nonlinear FAS for PDEs when the inequality constraints are removed (Appendix \ref{app:reductions}), PFAS reduces to linear multigrid.  The multilevel CD iteration of Tai \cite{Tai2003} is formulated in terms of an objective function, so as stated it only applies to optimization-type problems, and furthermore each of its coarser-level problems refers to the finest-level discretization.  The latter property is improved by reformulation using defect constraints and monotone injection operators (see \cite[Algorithm 4.7]{GraeserKornhuber2009}, or Section \ref{sec:cdmultilevel} herein).  The monotone multigrid methods of Kornhuber and co-authors \cite{GraeserKornhuber2009,Kornhuber1994} appear to be formulated only with Gauss--Seidel type smoothers, and applied only to optimization-type VI problems \cite[for example]{JouvetGraeser2013}.  Unlike the fastest variants of monotone multigrid, no zeroing of the basis functions associated with the active set (truncation) is employed in our method, both simplifying the implementation and allowing the coarse grid corrections to update the estimate of the active set on the fine grid. Numerical evidence indicates that the FAS construction of coarse grid corrections is useful for this purpose, even when the coarse grids are too coarse to usefully resolve the active set.
\pef{Insert comparison against the spiral problem once we have set rtol = 1e-12}
A limitation of FASCD is that, as formulated, it only solves bound-constrained VI problems.

@@

To construct box constraints on all meshes in the hierarchy, let $\tilde{\mathcal{V}}^j$ denote the set of functions on the vertices of $\mathcal{T}^j$ with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  The finest-level \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ are constructed by suitable interpolation of $\underline{\gamma}$, $\overline{\gamma}$ (e.g.~by applying the Cl\'ement interpolation operator~\cite{Carstensen2006}).  They must satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J. \label{eq:fe:boxconstraintrequirements}
\end{equation}
Analogously, we construct the discrete boundary data $g_D^J$ so that $\underline{\gamma}^J \le g_D^J \le \overline{\gamma}^J$ on those vertices of $\mathcal{T}^J$ which lie in $\partial_D \Omega$.  Then we define
\begin{equation}
\mathcal{K}^J = \left\{w\,:\,\underline{\gamma}^J \le w \le \overline{\gamma}^J \text{ on vertices of } \mathcal{T}^J, \, \text{ and } \, w|_{\partial_D\Omega} = g_D^J|_{\partial_D\Omega}\right\} \subset \mathcal{V}^J, \label{eq:fe:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of admissible FE functions. This is a nonempty, closed, and convex subset of $\mathcal{V}^J$.

@@

The monotone injections $\maxR,\minR$ \cite{Tai2003,GraeserKornhuber2009}, which will only be applied to constraints in $\tilde{\mathcal{V}}^j$, are nonlinear maps which maximize and minimize, respectively, nodal values over the open supports of coarser-mesh basis functions (i.e.~over the stars of the coarse vertices), Figure \ref{fig:Rplusminus}. These operators are constructed to enforce the property
\begin{equation}
\minR z \le z \le \maxR z,  \label{eq:fe:monotoneproperty}
\end{equation}
which does not hold for the standard injection operator $\iR$.
To make the construction precise, let $x_q^j$ denote the vertices of mesh $\mathcal{T}^j$, with corresponding basis functions $\psi_q^j(x)$.  For a $\tilde\RR$-valued nodal function $z\in\tilde{\mathcal{V}}^j$ we define functions in $\tilde{\mathcal{V}}^{j-1}$:
\begin{align}
(\maxR z)(x_p^{j-1}) &= \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}, \label{eq:fe:monotoneinjections} \\
(\minR z)(x_p^{j-1}) &= \min \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}. \notag
\end{align}
For standard refinement a coarse-mesh node $x_p^{j-1}$ is in the set of fine vertices $x_q^j$ such that $\psi_p^{j-1}(x_q^j)>0$, so \eqref{eq:fe:monotoneproperty} follows.  In the next Section we use $\maxR,\minR$ to construct ``level defect constraints'' such that the zero function is always admissible.  This construction depends on an additional ordering property of the monotone injections:
\begin{equation}
z\ge 0 \implies \minR z \ge 0, \qquad z \le 0 \implies \maxR z \le 0. \label{eq:fe:monotoneadditional}
\end{equation}

@@

\begin{example} \label{ex:results:classical}
Consider the unilateral 2D Laplacian obstacle problem, Example \ref{ex:plaplacian} with $p=2$ and admissible set $\mathcal{K} = \{v \ge \psi\} \subset W^{1,2}(\Omega)$.  We consider two problems, both over square domains $\Omega$, with the very different coincidence (active) sets $\{x\in\Omega \,:\, u(x)=\psi(x)\}$ shown in Figure \ref{fig:results:classical}.  The standard ``ball'' problem has an upper hemisphere obstacle $\psi$ \cite[Chapter 12]{Bueler2021}.  For the ``spiral'' problem \cite[problem 7.1.1]{GraeserKornhuber2009}, $\psi$ is constructed to produce the coincidence set shown.  For this experiment we used a coarsest regular-triangulation mesh with $m_0=5^2$ vertices (32 triangles).  The finest mesh, built by standard refinement, had $m_9=2049^2 \approx 4.2 \times 10^{6}$ vertices.  As our smoother we applied a single RS Newton iteration (Section \ref{sec:implementation}), with three (fixed) IC-preconditioned CG iterations to approximately solve the arising linear systems.  The coarse solver was RS$+$LU, iterated to convergence.  Tolerances \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$ were used in stopping criterion \eqref{eq:stoppingcriterion}.  The initial iterate was set to $w=\max\{0,\psi\}$.
\pef{For a fair comparison with the spiral problem in \cite{GraeserKornhuber2009}, I think we should set rtol = 1e-12}

@@

The 2D solver applied $P_1$ elements on a single core, but in 3D the elements were chosen to be $Q_1$ (hexahedra) on $P=16$ cores.  A coarse mesh with $m_0=16^d$ vertices minimally-resolved the coincidence set.  The smoother was two (fixed) RS Newton iterations, with three preconditioned GMRES iterations to approximately solve the arising linear systems; the preconditioning was additive Schwarz method (ASM) with incomplete factorization (ILU) on each process.  The coarse solver was RS Newton and LU, iterated to convergence.  Tolerances \id{rtol} $=10^{-5}$ and \id{atol} $=10^{-9}$ were used in \eqref{eq:stoppingcriterion}.

@@

@@

With these modifications, Algorithm \ref{alg:fascd} becomes the FAS V-cycle below, as a solver for \eqref{eq:app:fas:pde}. This algorithm appears in various sources in the literature, for example \cite[Algorithm 14]{Bruneetal2015} and \cite[section 5.3.4]{Trottenbergetal2001}.
